Demers Cartogram with Rivers

### 2024-07-02 paper/main.tex
Updates from Overleaf

Among the four types of cartogram categorized in a survey by \citet{nusrat2016State} (See \Cref{sec:RelatedWork} for definitions of contiguous, non-contiguous, rectangular, and Dorling), a trade-off is made between types of accuracy (See \Cref{table:accuracy}).
For this project, we focus on non-contiguous cartograms like Demers cartograms because they facilitate statistical comparison between regions, they can make good use of screen space, and comparison of regions is useful when studying Electronic Health Records (EHR) data.
Demers cartograms offer the advantage in cases where the data is not directly correlated to region sizes.
In other words, Demers cartograms are useful when a data dimension is not describing the geography of the region it represents but is tied to something else, for example the health of its population.
Also, the comparison of magnitudes becomes an area estimation task, which is effective for a numeric data encoding \cite{munzner2014Visualization}.
See \citea{nickel2022Multicriteria} for a more complete description of the advantages that Demers cartograms offer.
\new{One of the drawbacks of Demers' cartograms is that they may become more difficult to read when a region becomes displaced from its geospatial origin through the node layout process.
The layout of these more abstract shapes may simultaneously reduce the map's legibility and increase error.
See \citet{tong2018Cartograms} for a more detailed explanation.}
Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy.
Standard Demers cartograms are composed of square nodes that represent geographic enumeration units.
As such, this can reduce their legibility.
We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographic enumeration units, in this case, a Clinical Commissioning Group (CCG).
We hypothesize that introducing rivers improves the overall legibility of a cartogram.
By \textit{legibility} we mean readability and ability to interpret the cartogram.
To assess this hypothesis, we designed an experimental setup where participants engaged in correspondence and location tasks as part of a user study.
To reduce error and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout.
We then apply the algorithm to a real-world case study using EHR data to evaluate the result.
We present a user study that demonstrates its effectiveness.

@@

\bobgraph{Peer-reviewed Applications: } There has been a substantial amount of research done in this area, here we review some of the important work that has inspired our work.\citet{warf2008Geography} use a Dorling cartogram to represent religious diversity in the US.
\citet{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms.
\citet{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China.
\citet{tong2018Cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove the overlap of squares based on topological features, aiming to improve both geographical and topological accuracy.
\citet{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms using multiple data sets that include demographic, agriculture, and retail data in the US.
See \Cref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.

Our work extends the algorithm described by \citet{tong2018Cartograms} which incorporates a static topological feature into Demers cartograms.
Our work enhances that of \citet{tong2018Cartograms} in multiple ways.
First, we introduce multiple features (rivers) into the layout, as opposed to a single river.
Second, we make topological features dynamic and further improve legibility and geographical accuracy.
By the term \textit{dynamic}, we mean that the position of the rivers is updated as part of the layout algorithm.
In previous work, the river is static and serves merely as a boundary.
\new{To illustrate this more clearly, we refer to the video demonstration published by \citet{tong2018Cartograms}. Because the behavior of the layout algorithm is dynamic, a video is more appropriate to convey the motivation of dynamic features of interest.
We can observe the cartographic layout algorithm at 1:30 and 2:40 of the video.
During the update process, we can clearly observe nodes cross the Thames River as the size expands.
As the size of all the nodes expand, a number of nodes are pushed south in order to make more use of screen space.
If we want to make the most efficient use of screen space, we need to translate the Thames River further south during the layout process in order to prevent nodes from crossing it.
This is one of the main motivations for introducing dynamic rivers. }
Third, we improve the algorithm to resolve stalemates.
Finally, the way we evaluate the cartograms is also different.
\citet{tong2018Cartograms} count river crossings to evaluate error (a statistical metric).
Here our focus is on readability, and thus we include a user study.

\bobgraph{Cartograms in Media: }Cartograms are an engaging visual representation and therefore they are a popular choice of representation in covering various topics by the media.
The Washington Post uses cartograms to visualize the US overseas economic assistance, in arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}.
National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}.
\citet{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal.
\citet{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

\bobgraph{River Shapefiles:} We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for the Thames River, the Trent River, and the Great Ouse River in England.
These rivers were chosen as they are well-known rivers, pass through regions with dense populations, and provide informative geographical and topological cues.
Although including smaller rivers is technically feasible, it may not increase the legibility of the cartogram.
This is an open question for future work.

@@

\bobgraph{EHR Data: }We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}.
The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS.
Some datasets include:
\begin{itemize}
    \setlength\itemsep{0px}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

Each CCG has a unique ONS code, which is used to link the CCG shapefile with the statistical data.

@@

{
    \begin{figure*}[tb!]
        \centering
        \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{flowchart.png}
        \caption{An overview of our hybrid layout algorithm incorporating rivers.
            See also \Cref{alg:UpdateLayout} in Section \ref{sec:main} for more detail.
            See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate.
            For illustration purposes, we show the rendered views alongside the logical views representing the actual computation, \new{and set all squares to the same size for clarity.}}
        \label{fig:flowchart}
    \end{figure*}
}

@@

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average time to locate a CCG measured for four CCGs.
The chart indicates that the introduction of rivers has slightly reduced the time needed to locate CCGs.
The two groups show no statistically significant differences according to the two-sample unequal variance $t$-test performed, $t$(356.03) = 0.87, $p$ = 0.38.
However, the average time to locate a given CCG is reduced for 3 out of the 4 CCGs users in the study.

@@

\section{Obtaining Shapefile}
\label{obtaining-shapefile}
\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}
\section{Pre-processing Shapefiles}
\label{app:pre-processing}
Shapefiles from different sources are likely to be incompatible.
In our case, the NHS CCG shapefile is incompatible with the river shapefiles.
The major reason for the incompatibility is the coordinate reference system (CRS).
The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid).
The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System).
Here, we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility and reduce shapefile size to improve performance.

@@

    \begin{algorithm}[H]
        \caption{Procedure to derive a point based an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \EdgeStart $ \\

@@

    \begin{algorithm}[H]
        \caption{Procedure to derive a corridor to resolve stalemate.
We use an SVG canvas, where the point of origin (0,0) is located at the top left corner, with the x-axis extending to the right and the y-axis extending downwards (See \Cref{fig:stalemate}).}\label{alg:derive corridor}

>>>>

Among the four types of cartogram categorized in a survey by \citet{nusrat2016State} (See \Cref{sec:RelatedWork} for definitions of contiguous, non-contiguous, rectangular, and Dorling), a trade-off is made between types of accuracy (See \Cref{table:accuracy}).
For this project, we focus on non-contiguous cartograms like Demers cartograms because they facilitate statistical comparison between regions, they can make good use of screen space, and comparison of regions is useful when studying Electronic Health Records (EHR) data.
Demers cartograms offer the advantage in cases where the data is not directly correlated to region sizes.
In other words, Demers cartograms are useful when a data dimension is not describing the geography of the region it represents but is tied to something else, for example the health of its population.
Also, the comparison of magnitudes becomes an area estimation task, which is effective for a numeric data encoding \cite{munzner2014Visualization}.
See \citea{nickel2022Multicriteria} for a more complete description of the advantages that Demers cartograms offer.
\new{One of the drawbacks of Demers cartograms is that they may become more difficult to read when a region becomes displaced from its geospatial origin through the node layout process.
The layout of these more abstract shapes may simultaneously reduce the map's legibility and increase error.
See \citet{tong2018Cartograms} for a more detailed explanation.}
Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy.
Standard Demers cartograms are composed of square nodes that represent geographic enumeration units.
As such, this can reduce their legibility.
We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographic enumeration units, in this case, a Clinical Commissioning Group (CCG).
We hypothesize that introducing rivers improves the overall legibility of a cartogram.
By \textit{legibility} we mean readability and ability to interpret the cartogram.
To assess this hypothesis, we designed an experimental setup where participants engaged in correspondence and location tasks as part of a user study.
To reduce error and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout.
We then apply the algorithm to a real-world case study using EHR data to evaluate the result.
We present a user study that demonstrates its effectiveness.

@@

\bobgraph{Peer-reviewed Applications: } There has been a substantial amount of research done in this area, here we review some of the important work that has inspired our work. \citet{warf2008Geography} use a Dorling cartogram to represent religious diversity in the US.
\citet{sun2010Effectiveness} depict 1996 US election data and 2005 Chinese population data using Dorling, Mosaic, and contiguous cartograms.
\citet{gao2020Visualising} present a Dorling cartogram to illustrate COVID-19 infections in China.
\citet{tong2018Cartograms} use a Demers cartogram to visualize health-related data by CCG regions in England, the work introduces a novel technique to remove the overlap of squares based on topological features, aiming to improve both geographical and topological accuracy.
\citet{nusrat2020Recognition} investigate the memorability of contiguous and Dorling cartograms using multiple data sets that include demographic, agriculture, and retail data in the US.
See \Cref{table:region vs node} for a list of literature that adopts cartograms for visualization with corresponding geographical regions and node counts.

Our work extends the algorithm described by \citet{tong2018Cartograms} which incorporates a static topological feature into Demers cartograms.
Our work enhances that of \citet{tong2018Cartograms} in multiple ways.
First, we introduce multiple features (rivers) into the layout, as opposed to a single river.
Second, we make topological features dynamic and further improve legibility and geographical accuracy.
By the term \textit{dynamic}, we mean that the position of the rivers is updated as part of the layout algorithm.
In previous work, the river is static and serves merely as a boundary.
\new{To illustrate this more clearly, we refer to the video demonstration published by \citet{tong2018Cartograms}. Because the behavior of the layout algorithm is dynamic, a video is more appropriate to convey the motivation of dynamic features of interest.
We can observe the cartographic layout algorithm at 1:30 and 2:40 of the video.
During the update process, we can clearly observe nodes cross the River Thames as the size expands.
As the size of all the nodes expands, some nodes are pushed south to make more use of screen space.
If we want to make the most efficient use of screen space, we need to translate the River Thames further south during the layout process to prevent nodes from crossing it.
This is one of the main motivations for introducing dynamic rivers. }
Third, we improve the algorithm to resolve stalemates.
Finally, the way we evaluate the cartograms is also different.
\citet{tong2018Cartograms} count river crossings to evaluate error (a statistical metric).
Here our focus is on readability, and thus we include a user study.

\bobgraph{Cartograms in Media: }Cartograms are an engaging visual representation and therefore they are a popular choice of representation in covering various topics by the media.
The Washington Post uses cartograms to visualize the US overseas economic assistance, in arms sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}.
National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}.
\citet{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal.
\citet{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

\bobgraph{River Shapefiles:} We used OpenStreetMap \cite{openstreetmapRelation} as our data source to obtain shapefiles for the River Thames, the Trent River, and the Great Ouse River in England.
These rivers were chosen as they are well-known rivers, pass through regions with dense populations, and provide informative geographical and topological cues.
Although including smaller rivers is technically feasible, it may not increase the legibility of the cartogram.
This is an open question for future work.

@@

\bobgraph{EHR Data: }We obtained the Clinical Commissioning Group Outcomes Indicator Set (CCG OIS) from NHS Digital \cite{nhsdigitalClinical}.
\new{See \ref{dataset} for a detailed description of the datasets used.}
The OIS is a set of indicators that are used to measure the quality of care and the associated health outcomes in the NHS.
Each CCG has a unique ONS code, which is used to link the CCG shapefile with the statistical data.

@@

{
    \begin{figure*}[tb!]
        \centering
        \includegraphics[width=0.8\textwidth,height=\textheight,keepaspectratio]{flowchart.png}
        \caption{An overview of our hybrid layout algorithm incorporating rivers.
            See also \Cref{alg:UpdateLayout} in Section \ref{sec:main} for more detail.
            See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate.
            For illustration purposes, we show the rendered views alongside the logical views representing the actual computation, \new{and use the same size for all squares for clarity.}}
        \label{fig:flowchart}
    \end{figure*}
}

@@

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average time to locate a CCG measured for four CCGs.
The chart indicates that the introduction of rivers has slightly reduced the time needed to locate CCGs.
The two groups show no statistically significant differences according to the two-sample unequal variance $t$-test performed, $t$(356.03) = 0.87, $p$ = 0.38.
However, the average time to locate a given CCG is reduced for 3 out of the 4 CCGs in the study.

@@

\section{Dataset Description}
\label{dataset}
\begin{itemize}
    \setlength\itemsep{0px}
    \item Under 75 mortality: cardiovascular disease, respiratory disease, liver disease, and cancer
    \item Emergency hospital admission: stroke, alcohol-specific admission and readmission, coronary heart disease, re-admissions within 30 days of discharge, children with lower respiratory tract infections
\end{itemize}

\section{Obtaining Shapefiles}
\label{obtaining-shapefile}
\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}
\section{Pre-processing Shapefiles}
\label{app:pre-processing}
Shapefiles from different sources are likely to be incompatible.
In our case, the NHS CCG shapefile is incompatible with the river shapefiles.
The major reason for the incompatibility is the coordinate reference system (CRS).
The CRS of the CCG shapefile is EPSG:27700 (OSGB36 - British National Grid).
The CRS of the river shapefiles is EPSG:4326 (WGS84 - World Geodetic System).
Here, we provide some pre-processing steps using QGIS (version: 3.26.0-Buenos Aires) \cite{qgisWelcome} to handle the incompatibility and reduce shapefile size to improve performance.

@@

    \begin{algorithm}[H]
        \caption{Procedure to derive a point based on an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \EdgeStart $ \\

@@

    \begin{algorithm}[H]
        \caption{Procedure to derive a corridor to resolve stalemates.
We use an SVG canvas, where the point of origin (0,0) is located at the top left corner, with the x-axis extending to the right and the y-axis extending downwards (See \Cref{fig:stalemate}).}\label{alg:derive corridor}
